EDIpack2: interoperable Lanczos-based solver for generic quantum impurity problems

### 2025-05-11 02_install.tex
Updated draft
- use subfiles, so we can compile in every subfile
- written something re. umatrix
- updated gitignore
Finish polishing 03_edipack text Quick review of 02_install.

\section{Installation}\label{SecInstall}
The installation of \NAME is available through CMake which ensures
multi-platforms compabitility and dependencies resolution.  
The software builds into two distinct libraries.
The main one is {\tt  libedipack.a} which, alongside the generated Fortan
modules, wraps the \NAME software possibly including support for  
inequivalent impurities.
A second dynamic library, {\tt libedipack\_cbinding.so}, enables interoperability through
specific bindings to the C programming language.

\subsection{Structure}\label{sSecInstallStructure}
\NAME is a modular library which contains three principal structures.
At the core is the exact diagonalization solver: \NAME.
Next there is a {\tt EDIpack2ineq} which extends application to the case of multiple
inequivalent impurity problems. Finally, there we provide a Fortran-C/C++
interface, which enables development of additional API or
interoperability with other, external, software.

\begin{itemize}
\item{\bf EDIpack2.}
This constitutes the  building block of the whole software. This part
implement the  with the Lanczos-based solver for generic quantum
impurity systems encoding different symmetries, i.e. quantum number
conservations and apt to solve multi-orbital problems, also in
presence of coupling to local phonons.  
The \NAME solver has a hierarchical and modular structure: different
sections of the library communicate through a shared memory layer. The
top module of the library is {\tt EDIPACK2} which, once loaded,
enables access to the Fortran API in terms of suitable procedures to
initialize, execute and finalize the solver or to retrieve internal
quantities while making opaque to the user the internal structure of
the library. 
A detailed presentation of the library can be found in
Sec.\ref{SecEDIpack}.

\item{\bf EDIpack2ineq.}
This part of the software, leveraging on the object oriented concepts
available in modern Fortran, aims to extend the \NAME library to the
case of multiple inequivalent and independent impurity problems. This
is particularly useful while using \NAME as a solver for DMFT in
presence of unit cells with inequivalent atoms, for systems with
somehow broken translational symmetry (e.g. heterostructures, large
supercells, etc.).

\item{\bf EDIpack2 C-bindings.}
\NAME includes a single module implementing a Fortran-C/C++ interface of the main
library procedures. The module is developed around the implicit {\tt
  ISO\_C\_BINDING} capabilities of the most recent Fortran
distributions, which enable to translate Fortran procedures directly
to C. In order to overcome all the difficulties related to the internal
structure of the library,  we interfaced all and just the
procedures and the variables exposed to the user. 
This module aims to foster interoperability of \NAME  with different
third party softwares as well as to support development of additional
API.

% \subsection{Dependent software}
% As a concrete example illustrating the interoperability of the
% software we here illustrate the case of EDIpy2, i.e. the Python API for
% \NAME and present an interface layer to Triqs (Toolbox
% for Reasearch In Quantum Systems).

% \item{\bf EDIpy2.}
% This is a simple Python module which provides Python API to the
% \NAME Fortran library. This interface is built around the Python
% support to C-types, which allows to import the dynamic C-binding
% library generated upon building \NAME. The module contains a specific
% class, whose methods mirrors through  duck-typing all the available
% procedures of \NAME as well as it gives access to relevant shared
% control variables.

% \item{\bf EDIpack2Triqs.}
% This is a thin interface layer from \NAME to Triqs,
% built around the Python API of the \NAME library. The exact
% diagonalization solver is encapsulated in a dedicated class,
% containing the necessary methods to initialize and run a single
% instance of the solver. The interface also includes a specific class
% encompassing the effective discretized bath structure as well as their 
% optimization methods.

\end{itemize}

>>>>

\documentclass[edipack2.tex]{subfiles}
\begin{document}

\section{Installation}\label{SecInstall}
The installation of \NAME is available through CMake which ensures
multi-platforms compabitility and dependencies resolution.  
The software builds into two distinct libraries.
The main one is {\tt  libedipack.a} which, alongside the generated Fortan
modules, wraps the \NAME software possibly including support for  
inequivalent impurities.
A second dynamic library, {\tt libedipack\_cbinding.so}, enables interoperability through
specific bindings to the C/C++ programming language.

\subsection{Structure}\label{sSecInstallStructure}
\NAME is a modular library organized into three main components. At its 
core is the exact diagonalization solver, which provides the primary 
computational engine. Building on this, the library includes 
{\tt EDIpack2ineq}, an extension for handling multiple inequivalent 
impurity problems. Finally, a Fortran-C/C++ interface is provided for 
seamless integration with external software and the development of 
additional APIs.

\begin{itemize}
\item{\bf EDIpack2.}
  This module forms the foundation of the library, implementing the 
  Lanczos-based solver for general quantum impurity problems. It 
  supports systems with a wide range of symmetries, including quantum 
  number conservation, multi-orbital models, and even coupling to local 
  phonons. The \NAME solver is structured hierarchically and modularly, 
  with different sections of the library communicating through a shared 
  memory layer. The top-level module, {\tt EDIPACK2}, provides access 
  to the core Fortran API, exposing key procedures for initialization, 
  execution, and finalization, while abstracting the underlying data 
  structures. A detailed overview of this part of the library is 
  provided in Sec.\ref{SecEDIpack}.

\item{\bf EDIpack2ineq.}
  This extension, developed using modern Fortran constructs,
  enables the treatment of multiple, independent quantum 
impurities. It is particularly useful in DMFT applications involving 
unit cells with inequivalent atomic sites or systems with broken 
translational symmetry, such as heterostructures, large supercells, 
or disordered materials. This module provides flexible memory 
management and supports the simultaneous solution of multiple impurity 
problems, as discussed in Sec.\ref{sSecIneq}.

\item{\bf EDIpack2 C-bindings.}
  For enhanced interoperability, \NAME includes a dedicated module 
implementing Fortran-C/C++ bindings for key library procedures. This 
module relies on the {\tt ISO\_C\_BINDING} capabilities available in 
modern Fortran, allowing direct translation of Fortran functions to C. 
To ensure a straightforward user experience, only the functions and 
variables directly exposed to the user are included in this binding, 
shielding developers from the complexity of the library's internal 
architecture. This interface is intended to facilitate integration 
with third-party software and support the development of custom APIs.
\end{itemize}

### 2025-05-11 03_edipack2.tex
Finish polishing 03_edipack text Quick review of 02_install.

\subsection{EDIpack2ineq: inequivalent impurities}\label{sSecIneq}
It is often necessary to solve systems with several, independent,
quantum impurities. Paradigmatic examples arise in the DMFT
framework when dealing with the effective description of lattices
built out of unit cells hosting multiple inequivalent atoms or large
super-cell systems with possibly broken translational symmetries,
e.g. heterostructures of correlated materials, disordered systems,
etc.
As the structure of \NAME allows the presence of a single instance of
the solver at any time, we implemented a suitable extension to deal
with multiple impurities. Specifically {\tt EDIpack2Ineq} is a
sub-library which complements \NAME, wrapping memory and procedures in
order to enable simultaneous treatment of  inequivalent impurities. 
Using standard Fortran interfaces it is possible to group all the
\NAME and {\tt EDIpack2ineq} functions under the same name without
compromising the software functionalities.  
Here below we briefly discuss the main additions introduced by {\tt
  EDIpack2ineq}.

\subsubsection{Structure}\label{ssSecIneqStructure}
The sub-library is composed of a number of Fortran modules which are
wrapped by {\tt EDIPACK2INEQ}. This latter represents the main user
interface and gives access to all the available procedures and
variables as needed to solve quantum impurity problems.
The user needs to invoke use of this module alongside the \NAME one:
\begin{lstlisting}[style=fstyle,numbers=none,basicstyle={\scriptsize\ttfamily}]
program test
  !Load EDIpack2 library 
  USE EDIPACK2 
  !Load the Inequivalent impurities extension
  USE EDIPACK2INEQ
  ...
\end{lstlisting}

A major part of the {\tt EDIpack2ineq} library is to define global
variables which wrap and conserve the memory pool of \NAME.
For instance we implement a number of
globally shared arrays with an increased rank to store the
variables in any memory instance of \NAME and accessible through
the input/output procedures (see \secu{sSecIO}).
Specific objects, e.g. {\tt effective\_bath} or {\tt gfmatrix}, in
principle require dedicated MPI
communication when parallel execution of the solver among independent
impurities. However, we opt for the much simpler and economic solution
of reading these objects from the file in which they get saved
anyway.

\subsubsection{Core routines}\label{ssSecIneqGlobal}
The module {\tt E2I\_MAIN} wraps the extensions to the main algorithms
of \NAME, namely: initialization, diagonalization and finalization,
preserving the original naming convention.  
The initialization process is handled by {\tt ed\_init} which is
extended to accepts a bath array or rank-2 (as opposed to rank-1 array
in the \NAME implementation). The leading dimension now determines the
number of inequivalent impurity problems to be solved. This procedure
initialize all the {\tt EDIpack2ineq} variables and prepare each bath
for the diagonalization step. 
In {\tt ed\_solve}, which accepts the same rank-2 bath array, performs the
diagonalization of each quantum impurity problem in either serial of
parallel mode. The parallel mode is further controlled by the input
function {\tt mpi\_lanc=T/F}. If true Lanczos diagonalization is
performed in parallel and inequivalent sites are treated
sequentially. The inverse if false.   
The finalization procedure with full memory release is performed in
{\tt ed\_finalize}.

\subsubsection{Inequivalent Baths}\label{ssSecIneqBath}
The inequivalent impurities extension is encoded in {\tt E2I\_BATH}
which contains procedures to control the baths setup on the user side
and the  implementation of conventional symmetry operations.
In the {\tt E2I\_BATH\_REPLICA} module we extend all the functions
which set the replica/general bath matrix basis 
$\vec{\Gamma}$ and the corresponding initial variational parameters
$\vec{\lambda}$. Although the variational parameters are generally
different from site to site, in the current implementation the matrix
basis is independent from the impurities number. This will possibly be
relaxed in a future release. 
Analogously, in the module {\tt E2I\_BATH\_FIT} we implement an extension of the generic
function {\tt ed\_chi2\_fitgf} to perform multiple independent
bath optimization for all inequivalent impurities in terms of an MPI iterative cycle.

\subsubsection{Input/Output}\label{ssSecIneqIO}
A key part of the {\tt EDIpack2ineq} library is the extension for the
input/output procedures to return observables or dynamical functions per inequivalent impurity.
The module {\tt E2I\_IO} implements numerous functions which retrieve
or evaluate on-the-fly any quantity accessible in \NAME for any given
impurity (or all of them) using the same naming original convention
defined in \NAME.

>>>>

\subsection{EDIpack2ineq: inequivalent impurities}\label{sSecIneq}
In many contexts, it is necessary to solve systems with multiple, 
independent quantum impurities. This situation frequently arises in 
DMFT when modeling lattices with complex unit cells containing 
inequivalent atomic sites, or in supercell calculations where 
translational symmetry is broken, as in heterostructures, disordered 
systems, or multiorbital setups.

Given that \NAME allows only a single instance of the solver at a 
time, a dedicated extension is required to handle these more complex 
cases. To address this, we developed the {\tt EDIpack2ineq} 
sub-library, which extends the base functionality of \NAME by managing 
memory and procedures for multiple impurities. This approach ensures 
that all functions remain accessible through standard Fortran 
interfaces, seamlessly integrating both \NAME and {\tt EDIpack2ineq} 
without compromising overall software design.

Below, we provide a brief overview of the main features introduced by 
{\tt EDIpack2ineq}.

\subsubsection{Structure}\label{ssSecIneqStructure}
The core of the {\tt EDIpack2ineq} sub-library consists of several 
Fortran modules, all wrapped under the main interface module 
{\tt EDIPACK2INEQ}. This module provides access to the full range of 
procedures and variables required to solve inequivalent quantum 
impurity problems. To use this extension, the user must include both 
the main \NAME and the {\tt EDIpack2ineq} modules, as shown below:
\begin{lstlisting}[style=fstyle,numbers=none,basicstyle={\scriptsize\ttfamily}]
program test
  !Load EDIpack2 library 
  USE EDIPACK2 
  !Load the Inequivalent impurities extension
  USE EDIPACK2INEQ
  ...
\end{lstlisting}

A significant part of the {\tt EDIpack2ineq} extension is the 
definition of global variables that extend the memory pool of 
\NAME. This includes higher-rank arrays for storing impurity-specific 
data, ensuring that each impurity's state is maintained separately 
during the computation. While dedicated MPI communication could be 
used to manage specific objects in parallel, e.g. {\tt
  effective\_bath} or {\tt gfmatrix}, we opted for a simpler and safer  
file-based approach to maintain compatibility with the standard I/O 
procedures described in \secu{sSecIO}.

\subsubsection{Core routines}\label{ssSecIneqGlobal}
The module {\tt E2I\_MAIN} wraps the key extensions to the main 
algorithms of \NAME, including initialization, diagonalization, and 
finalization, while preserving the original function names for 
consistency.

\paragraph{\bf Initialization} The function {\tt ed\_init} is extended to 
  accept a rank-2 bath array, where the leading dimension specifies 
  the number of inequivalent impurity problems. This ensures each bath 
  is properly initialized for subsequent diagonalization.

\paragraph{\bf Diagonalization} The function {\tt ed\_solve} also accepts this 
  rank-2 bath array and manages the diagonalization of each impurity 
  problem. Parallel execution is controlled by the input flag 
  {\tt mpi\_lanc=T/F}, which determines whether the Lanczos 
  diagonalization is performed in parallel or sequentially across 
  different impurities.

\paragraph{\bf Finalization} Memory release is handled through the extended 
{\tt ed\_finalize} function, which clears the memory pool for all 
inequivalent impurity instances.

\subsubsection{Inequivalent Baths}\label{ssSecIneqBath}
The bath setup for multiple impurities is managed by the 
{\tt E2I\_BATH} module, which provides functions for defining 
site-specific bath matrices and their corresponding variational 
parameters. This includes support for conventional symmetry operations 
and the handling of replica bath structures.

In the module {\tt E2I\_BATH\_REPLICA}, we extend the 
matrix basis definition for the variational parameters 
$\vec{\lambda}$, allowing for flexible bath optimization across 
multiple sites. Although the matrix basis is currently shared among 
all impurities, future versions may allow for fully independent 
bath parameterizations.

Additionally, the {\tt E2I\_BATH\_FIT} module extends the generic 
{\tt ed\_chi2\_fitgf} function to support simultaneous, independent 
bath optimization for all impurities using MPI, thereby improving the 
efficiency of large-scale calculations.

\subsubsection{Input/Output}\label{ssSecIneqIO}
A key component of the {\tt EDIpack2ineq} extension is the 
enhanced I/O capability for handling impurity-specific observables. 
The {\tt E2I\_IO} module includes a variety of functions for 
retrieving site-resolved quantities, such as local Green's functions, 
self-energies, and density matrices. These functions use the same 
naming conventions as the core \NAME library, ensuring a consistent 
user experience across single and multi-impurity calculations.
